---
title: "Class 05: Data Visualization with GGPLOT"
author: "Alisa Zhang (PID: A18299618)"
format: gfm
---

Today we are exploring the **ggplot** package and how to make nice figures in R.

There are lots of ways to make figures and plots in R. These include:

-   so called "base" R
-   and add on packages like **ggplot2**

Here is a simple "base" R plot.

```{r}
head(cars)
```

We can simply pass this to the `plot()` functions.

```{r}
plot(cars)
```

> Key-point: Base R is quick but simple looking.

Let's see how we can plot this with the **ggplot2**...

First, we need to install this add-on package. For this we can use the `install.packages()` function - WE DO THIS IN THE CONSOLE, NOT our report.

Second, we need to load the package with the `library()` function everytime we want to use it.

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is composed of at least 3 layers:

-   **data** (i.e. a data.frame with teh things you want to plot),
-   aesthetics **aes()** that map the columns of data to your plot features
-   geoms like **geom_point()** that shows how the plot appears

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point()
```

> For simple "canned" graphs base R is quicker, but as things get more custom and elobrate then ggplot wins out.

Lets add more layers to out ggplot

Add a line showing the relationship between x and y Add a title Add custom axis labels "Speed (MPH)" and "Distance (ft)" Change the theme...

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  labs(
    title = "Speed vs Distance Plot",
    x = "Speed (MPH)",
    y = "Distance (ft)"
  ) +
  theme_light()
```

## Going further - Gene Expression Dataset

Read some gene expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this dataset

```{r}
nrow(genes)
```

```{r}
ncol(genes)
colnames(genes)
```

> Q2. How many upregulated genes are there

A useful function for counting up occurancee of things in a vector is the `table()` function.

```{r}
table(genes$State)
```


Make a v1 figure

```{r}
p <- ggplot(genes) +
  aes(
    x = Condition1,
    y = Condition2,
    colour = State
  ) +
  geom_point()
```

```{r}
p + 
  scale_color_manual(values = c("blue", "grey", "red")) +
  labs(
    title = "Gene Expression Changes Upon Drug Treatment",
    subtitle = "another scatterplot made with ggplot",
    x = "Control (No Drug)",
    y = "Drug Treatment"
  ) +
  theme_light()
```

## More Plotting Examples - `gapminder` Dataset

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```

```{r}
tail(gapminder, 3)
```

> Q4. How many different countries are in this dataset

```{r}
length(table(gapminder$country))
```

> Q5. Number of continents

```{r}
unique(gapminder$continent)
```

Version 1 plot

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, colour = continent) +
  geom_point()
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, colour = continent, label = country) +
  geom_point() +
  geom_text()
```

I can use the `ggrepel` package to make more sensible labels here

```{r}
library(ggrepel)

ggplot(gapminder) +
  aes(gdpPercap, lifeExp, colour = continent, label = country) +
  geom_point() +
  geom_text_repel()
```

I want a separate pannel per continent 

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, colour = continent, label = country) +
  geom_point() +
  facet_wrap(~continent)
```


Let’s focus on the main advantages of ggplot over base R:

1. **Layered Grammar of Graphics**: ggplot uses a consistent, layered approach. You build plots by adding layers (data, aesthetics, geoms, themes) with the `+` operator. This makes complex plots easier to construct and modify, compared to base R where each plot type often requires a different function and many arguments [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

2. **Publication-Quality Defaults**: ggplot produces attractive, publication-ready figures by default. Base R plots are quick but often require a lot of manual tweaking to look polished [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

3. **Declarative Syntax**: In ggplot, you declare what you want to show (e.g., which variables map to which aesthetics), rather than specifying every detail step-by-step as in base R [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

4. **Consistency and Reproducibility**: ggplot’s approach is consistent across plot types, making it easier to learn and reproduce complex visualizations. Base R can be more idiosyncratic, with different functions and arguments for each plot type [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

5. **Customization and Extensibility**: ggplot makes it easy to add layers (like trend lines, color scales, themes) and customize every aspect of the plot. Base R allows pixel-level control but is more cumbersome for complex customizations [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

