---
title: "Class 06: R Functions Lab"
author: "Alisa Zhang (PID: A18299618)"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

This week we are introducing *R Functions* and how to write our own R
functions.

Questions to answer:

> Q1. Write a function grade() to determine an overall grade from a
> vector of student homework assignment scores dropping the lowest
> single score. If a student misses a homework (i.e. has an NA value)
> this can be used as a score to be potentially dropped. Your final
> function should be adquately explained with code comments and be able
> to work on an example class gradebook such as this one in CSV format:
> “https://tinyurl.com/gradeinput” \[3pts\]

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

Follow the guidelines from class

-   Write a working snippet of code that solves a simple problem

```{r}
# mean()
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

mean(student1)
```

BUT we need to drop the lowest score. First we need to identify the
lowest score.

```{r}
# Which element of the vector is the lowest
which.min(student1)
```

What i want to do is to *drop* (i.e. exclude) this lowest score from my
mean() calculation.

```{r}
# This will return everything but the 8th element in the vector
student1[-8]
```

Now we can use the answer from which.min() to return all other elements
of the vector.

```{r}
# This is our first working snippet
mean( student1[-which.min(student1)] )
```

What about other example students? Will this work?

We could try using the na.rm=TRUE argument for mean but this is unfair.

```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
mean(student2, na.rm=TRUE)

student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
mean(student3, na.rm=TRUE)
```

Another approach is to mask (i.e. replace) all NA values with 0s.

First we need to find the NA elements of the elements.

```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
x <- student2

is.na(x)
which( is.na(x) )
```

Now we have identified the NA elements we want to "mask" them. How to
replace them with zero?

```{r}
x[is.na(x)] <- 0
x
mean(x)
```

Recall that we should drop the lowest score...

```{r}
x[is.na(x)] <- 0
mean( x[-which.min(x)] )
```

Now we are here with our code:)

```{r}
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
x <- student3
x[is.na(x)] <- 0
mean( x[-which.min(x)] )
```

## Now we make our function

Take the snippet and turn it into a function

Recall that every function has three parts: - A name, in our case
`grade()` - Input arguments, a vector of student scores - The body i.e.
our working snippet of code

Using Studio I will select 'Code \> Extract Function'

```{r}
grade <- function(x) {
  x[is.na(x)] <- 0
  mean( x[-which.min(x)] )
}
```

```{r}
grade(student1)
grade(student2)
grade(student3)
```

This looks great! We now need to add comments to explain this to our
future selves and others who want to use this function.

```{r}
#' Calculate the average score for a vector of a student's score dropping 
#' the lowest. Missing value would be replaced with zero.
#'
#' @param x A numeric vector of hw scores
#'
#' @returns Average score
#' @export
#'
#' @examples
#'  student <- c(100, NA, 90, 97)
#'  grade(student)
#'  

grade <- function(x) {
  # Mask NA with zeros
  # Treat missing values as zeros
  x[is.na(x)] <- 0
  # Exclude lowest score from mean
  mean( x[-which.min(x)] )
}
```

Now finally we can use our function on real class data from this CSV
format: “https://tinyurl.com/gradeinput”

```{r}
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names = 1)
```

```{r}
apply(gradebook, 1, grade)
```

> Q2. Using your grade() function and the supplied gradebook, Who is the
> top scoring student overall in the gradebook? \[3pts\]

To answer question2, we can run the apply() function and save the
result.

```{r}
result <- apply(gradebook, 1, grade)
which.max(result)
```

> Q3. From your analysis of the gradebook, which homework was toughest
> on students (i.e. obtained the lowest scores overall? \[2pts\]

```{r}
ave.score <- apply(gradebook, 2, mean, na.rm=TRUE)
ave.score
which.min(ave.score)
```

```{r}
med.score <- apply(gradebook, 2, median, na.rm=TRUE)
med.score
which.min(med.score)
```

```{r}
boxplot(gradebook)
```

So homework 2 is probably the lowest.

> Q4. Optional Extension: From your analysis of the gradebook, which
> homework was most predictive of overall score (i.e. highest
> correlation with average grade score)? \[1pt\]

Are the final result (i.e. average score of each student) correlated
with the result (i.e. score) of individual hws - the gradebook columns

```{r}
masked.gradebook <- gradebook
masked.gradebook[ is.na(masked.gradebook) ] <- 0
masked.gradebook
```

```{r}
cor(result, masked.gradebook$hw5)
```

```{r}
cor.hw <- apply(masked.gradebook, 2, cor, x=result)
cor.hw
which.max(cor.hw)
```

> Q5. Make sure you save your Quarto document and can click the “Render”
> (or Rmark- down”Knit”) button to generate a PDF foramt report without
> errors. Finally, submit your PDF to gradescope. \[1pt\]
